---
title: "Simple usage example"
author: "Kim Philipp Jablonski, Martin Pirkl"
date: "`r Sys.Date()`"
graphics: yes
output: BiocStyle::html_document
bibliography: bibliography.bib
vignette: >
    %\VignetteIndexEntry{Simple usage example}
    %\VignetteEngine{knitr::knitr}
    %\VignetteEncoding{UTF-8}
---

# Introduction

## Load required packages
```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  message=FALSE, out.width="\\textwidth", fig.align="center",
  strip.white=TRUE, warning=FALSE, tidy=TRUE,
  fig.height = 4, fig.width = 8, error=FALSE
)
```


Load the package with the library function.

```{r}
# fix "object 'guide_edge_colourbar' of mode 'function' was not found" when building vignettes (see also https://github.com/thomasp85/ggraph/issues/75)
library(ggraph)

library(tidyverse)
library(cowplot)
library(dce)

set.seed(42)
```

# Simple example

First we generate a random wild type network and then resample
its edge weights to generate the mutant network.

```{r}
node_num <- 10
edge_prob <- .5

graph_wt <- create_random_DAG(node_num, edge_prob)
graph_mt <- resample_edge_weights(graph_wt)

cowplot::plot_grid(
  plot_network(as(graph_wt, "matrix")),
  plot_network(as(graph_mt, "matrix"))
)
```

We can then simulate data for these two networks.

```{r}
X_wt <- simulate_data(graph_wt)
X_mt <- simulate_data(graph_mt)
```

Finally, it is possible to compute differential causal effects.

```{r}
res <- dce(graph_wt, X_wt, X_mt, solver = "lm")

res %>%
  as.data.frame %>%
  drop_na
```

The result can be visualized.

```{r}
plot(res)
```

# Session information

```{r}
sessionInfo()
```

# References
