---
title: "Simulations: Differential causal effects for pathway enrichment \n
Simulations of differential causal effects between two groups."
author: "Kim Jablonski, Martin Pirkl"
date: "`r Sys.Date()`"
graphics: yes
output: BiocStyle::pdf_document
vignette: >
    %\VignetteIndexEntry{dce}
    %\VignetteEngine{knitr::knitr}
    %\VignetteEncoding{UTF-8}
---

# Introduction



# Installation and loading
```{r global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, out.width="\\textwidth", fig.align="center",
                      strip.white=TRUE, warning=FALSE, tidy=TRUE,
                      #out.extra='style="display:block; margin:auto;"',
                      fig.height = 4, fig.width = 8, error=FALSE)
```

Install the package with the Bioconductor manager package.

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("dce")
```

Load the package with the library function.

```{r}
library(dce)

library(tidyverse)
library(purrr)

library(graph)
library(pcalg)
library(assertthat)


set.seed(42)
```

# Simple example

First we generate a random wild type network and then resample its edge weights to generate the mutant network.

```{r}
node.num <- 30
edge.prob <- .2

negweight.range <- c(-1,-.1)
posweight.range <- c(.1, 1)

graph.wt <- randomDAG(node.num, edge.prob, negweight.range, posweight.range)
graph.mt <- newWeights(graph.wt, negweight.range, posweight.range)
```

We can then simulate data on these two networks.

```{r}
X.wt <- rmvDAG(100, graph.wt)
X.mt <- rmvDAG(100, graph.mt)
```

Finally, it is possible to compute differential causal effects.

```{r}
res <- compute_differential_causal_effects(
    graph.wt, X.wt,
    graph.mt, X.mt
)

print(res)
```

The result can be visualized.

```{r}
plot(res)
```

# Session information

```{r}
sessionInfo()
```

# References:

* Taruttis et al
* Matthuis et al
* ...
